\chapter{Presentation of the four cases}
\label{chapter:presentation}

Using the process outlined in the methodology, four cases was chosen --- three successful and one unsuccessful --- from an initial list of 49. This chapter presents the empirical findings for each of these cases, starting with a summary in \tableref{presentation:summary} of the rationale for choosing each case.

\begin{margintable}[h]
  \begin{tabularx}{\textwidth}{ l X }
    \toprule
    Name       & Rationale for selection \\
    \midrule
    LogMeIn    & Publicly traded with a market capitalization of \$1 billion \abbr{USD}, thus a highly successful freemium-based company. The only among the four using an alternative product strategy. \\
    \midrule
    Evernote   & Have been very public about their numbers and has an exponential revenue growth. Have no network effects, but at the same time don't spent money on advertising. Very strong word-of-mouth effects. \\
    \midrule
    MailChimp  & Had been a premium-only company for eighth years when they changed to freemium, and experienced a tremendous growth both in users and profit. \\
    \midrule
    Helpstream & One of few companies that have been very public about their failure with freemium. More business-to-business oriented than the other case companies. \\
    \bottomrule
  \end{tabularx}
  \caption{The four cases, including a rationale for why each was chosen.}
  \label{table:presentation:summary}
\end{margintable}

\input{presentation.logmein}
\input{presentation.evernote}
\input{presentation.mailchimp}
\input{presentation.helpstream}